---
title: "Tweedieverse"
author: "Himel Mallick, Ali Rahnavard"
date: "`r BiocStyle::doc_date()`"
package: "`r if (requireNamespace('Tweedieverse', quietly = TRUE)) BiocStyle::pkg_ver('Tweedieverse') else 'Tweedieverse development version'`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Tweedieverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette shows compact, package-contained examples for three common
Tweedieverse workflow families:

- microbiome relative abundances,
- single-cell RNA-seq UMI counts,
- single-cell RNA-seq non-UMI/read counts.

The packaged files are small tab-delimited example excerpts stored in
`inst/extdata` so that the vignette can be built without internet access. They
are intended to exercise realistic input shapes, not to reproduce a full
biological analysis.

This vignette does **not** include simulation examples.

```{r packages}
suppressPackageStartupMessages({
  library(Tweedieverse)
  library(SummarizedExperiment)
  library(SingleCellExperiment)
  library(TreeSummarizedExperiment)
})
```

## Loading Packaged Example Data

Bioconductor package vignettes should run without downloading external files.
The helper below finds the packaged `inst/extdata` files both when the package
is installed and when the vignette is built from the source tree.

```{r data-helpers}
extdata_file <- function(file) {
  candidates <- c(
    system.file("extdata", file, package = "Tweedieverse"),
    file.path("inst", "extdata", file),
    file.path("..", "inst", "extdata", file)
  )
  candidates <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (length(candidates) == 0L) {
    stop("Could not find packaged extdata file: ", file)
  }
  candidates[[1]]
}

read_feature_table <- function(file) {
  tab <- utils::read.delim(
    extdata_file(file),
    row.names = 1,
    check.names = FALSE
  )
  as.matrix(tab)
}

read_metadata_table <- function(file) {
  utils::read.delim(
    extdata_file(file),
    row.names = 1,
    check.names = FALSE
  )
}

align_feature_metadata <- function(features, metadata) {
  shared <- intersect(colnames(features), rownames(metadata))
  if (length(shared) == 0L) {
    stop("No overlapping sample/cell IDs between feature columns and metadata rows.")
  }
  list(
    features = features[, shared, drop = FALSE],
    metadata = metadata[shared, , drop = FALSE]
  )
}
```

## Microbiome Relative Abundance Example

For microbiome data, `Tweedieverse()` accepts a
`TreeSummarizedExperiment`. The assay contains taxa/features in rows and
samples in columns; `colData` contains sample metadata.

```{r microbiome-container}
microbiome_features <- read_feature_table(
  "tweedieverse_microbiome_relative_abundance.tsv"
)
microbiome_metadata <- read_metadata_table(
  "tweedieverse_microbiome_metadata.tsv"
)

microbiome_aligned <- align_feature_metadata(
  microbiome_features,
  microbiome_metadata
)

microbiome_tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = microbiome_aligned$features),
  colData = microbiome_aligned$metadata
)

microbiome_tse
```

The example table contains relative abundances, so this analysis disables
additional offset normalization by setting `normalization = "NONE"`. For raw
microbiome counts, omit that argument to use the microbiome default `TSS`
size-factor offset.

```{r microbiome-fit}
microbiome_fit <- Tweedieverse(
  input_features = microbiome_tse,
  output = NULL,
  assay_name = "counts",
  domain = "microbiome",
  normalization = "NONE",
  fixed_effects = c("diagnosis", "antibiotics"),
  tweedie_p = NULL,
  median_comparison = FALSE,
  cores = 1
)

head(microbiome_fit)
```

## Single-Cell RNA-seq UMI Count Example

For single-cell RNA-seq data, `Tweedieverse()` accepts a
`SingleCellExperiment`. The UMI example uses integer molecule counts.

```{r umi-container}
umi_features <- read_feature_table("tweedieverse_scrna_umi_counts.tsv")
umi_metadata <- read_metadata_table("tweedieverse_scrna_umi_metadata.tsv")

umi_aligned <- align_feature_metadata(umi_features, umi_metadata)

umi_sce <- SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = umi_aligned$features),
  colData = umi_aligned$metadata
)

umi_sce
```

UMI counts are count-like. Here we pin the variance power to `p = 1`, the
Poisson variance case, and use the single-cell domain. `normalization = "SCRAN"`
computes size factors for the model offset without modifying the assay.

```{r umi-fit}
umi_fit <- Tweedieverse(
  input_features = umi_sce,
  output = NULL,
  assay_name = "counts",
  domain = "single_cell",
  normalization = "SCRAN",
  fixed_effects = c("condition", "batch"),
  tweedie_p = 1,
  median_comparison = FALSE,
  cores = 1
)

head(umi_fit)
```

## Single-Cell RNA-seq Non-UMI / Read Count Example

The non-UMI example uses read-count-like measurements. These can have a more
flexible mean-variance relationship than UMI counts, so we leave
`tweedie_p = NULL` and let Tweedieverse estimate the index after filtering.

```{r nonumi-container}
nonumi_features <- read_feature_table("tweedieverse_scrna_nonumi_counts.tsv")
nonumi_metadata <- read_metadata_table("tweedieverse_scrna_nonumi_metadata.tsv")

nonumi_aligned <- align_feature_metadata(nonumi_features, nonumi_metadata)

nonumi_sce <- SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = nonumi_aligned$features),
  colData = nonumi_aligned$metadata
)

nonumi_sce
```

```{r nonumi-fit}
nonumi_fit <- Tweedieverse(
  input_features = nonumi_sce,
  output = NULL,
  assay_name = "counts",
  domain = "single_cell",
  normalization = "SCRAN",
  fixed_effects = c("condition", "batch"),
  tweedie_p = NULL,
  median_comparison = FALSE,
  cores = 1
)

head(nonumi_fit)
```

## Multi-omics Container

The same containers can be collected into a `MultiAssayExperiment`. Each omics
layer is analyzed independently, and per-omics parameters can be supplied as
named vectors or lists.

```{r multiassay}
suppressPackageStartupMessages(library(MultiAssayExperiment))

tweedieverse_mae <- MultiAssayExperiment::MultiAssayExperiment(
  experiments = list(
    microbiome = microbiome_tse,
    scrna_umi = umi_sce,
    scrna_nonumi = nonumi_sce
  )
)

multi_fit <- Tweedieverse(
  input_features = tweedieverse_mae,
  output = NULL,
  domain = c(
    microbiome = "microbiome",
    scrna_umi = "single_cell",
    scrna_nonumi = "single_cell"
  ),
  normalization = c(
    microbiome = "NONE",
    scrna_umi = "SCRAN",
    scrna_nonumi = "SCRAN"
  ),
  fixed_effects = list(
    microbiome = c("diagnosis", "antibiotics"),
    scrna_umi = c("condition", "batch"),
    scrna_nonumi = c("condition", "batch")
  ),
  tweedie_p = list(
    microbiome = NULL,
    scrna_umi = 1,
    scrna_nonumi = NULL
  ),
  median_comparison = FALSE,
  cores = 1
)

names(multi_fit)
head(multi_fit$microbiome)
head(multi_fit$scrna_umi)
head(multi_fit$scrna_nonumi)
```

## Session Info

```{r session-info}
sessionInfo()
```
