---
title: "Tweedieverse Vignette"
author: "Himel Mallick, Ali Rahnavard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Tweedieverse Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(Tweedieverse)
```

## Overview

`Tweedieverse` fits per-feature Tweedie-based models for differential analysis of omics data.
This vignette provides:

- A minimal runnable toy example.
- An iHMP-style workflow example (not executed during vignette build).

## Minimal Runnable Example

```{r minimal-example}
set.seed(123)
features <- as.data.frame(matrix(rpois(24, lambda = 5), nrow = 12, ncol = 2))
colnames(features) <- c("feature1", "feature2")
rownames(features) <- paste0("sample", seq_len(nrow(features)))

metadata <- data.frame(
  group = rep(c("A", "B"), each = 6),
  row.names = rownames(features)
)

fit <- Tweedieverse(
  input_features = features,
  input_metadata = metadata,
  output = NULL,
  fixed_effects = "group",
  median_comparison = FALSE,
  cores = 1
)

head(fit)
```

## iHMP-Style Workflow Example

This example uses publicly available files that are commonly used in iHMP/HMP2-style workflows.
It is shown for reproducibility but is not executed during vignette build.

```{r ihmp-workflow, eval=FALSE}
library(data.table)
library(tibble)

input_features <- fread(
  "https://raw.githubusercontent.com/biobakery/Maaslin2/master/inst/extdata/HMP2_taxonomy.tsv",
  sep = "\t"
)
input_metadata <- fread(
  "https://raw.githubusercontent.com/biobakery/Maaslin2/master/inst/extdata/HMP2_metadata.tsv",
  sep = "\t"
)

features <- column_to_rownames(input_features, "ID")
metadata <- column_to_rownames(input_metadata, "ID")

hmp2_fit <- Tweedieverse(
  input_features = features,
  input_metadata = metadata,
  output = "./demo_output/HMP2",
  fixed_effects = c(
    "diagnosis",
    "dysbiosisnonIBD",
    "dysbiosisUC",
    "dysbiosisCD",
    "antibiotics",
    "age"
  ),
  random_effects = c("site", "subject"),
  base_model = "CPLM",
  adjust_offset = FALSE,
  cores = 8,
  median_comparison = TRUE,
  median_subtraction = TRUE,
  standardize = FALSE,
  reference = c("diagnosis,nonIBD")
)

head(hmp2_fit)
```

## Session Info

```{r session-info}
sessionInfo()
```
