---
title: "Tweedieverse"
author: "Himel Mallick, Ali Rahnavard"
date: "`r BiocStyle::doc_date()`"
package: "`r if (requireNamespace('Tweedieverse', quietly = TRUE)) BiocStyle::pkg_ver('Tweedieverse') else 'Tweedieverse development version'`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Tweedieverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`Tweedieverse` provides a unified differential analysis workflow for omics
measurements whose empirical variance structure is well represented by the
Tweedie family. The framework is intended for heterogeneous high-throughput
data, including microbiome abundances, single-cell RNA-seq measurements, bulk
RNA-seq summaries, metabolite intensities, and other continuous or count-like
assays.

The package uses Bioconductor containers as the primary interface. A single
omics assay is represented as a `SummarizedExperiment`,
`TreeSummarizedExperiment`, or `SingleCellExperiment`, and matched multi-omics
assays are represented as a `MultiAssayExperiment`. This keeps feature assays,
sample metadata, and experiment-level annotations together and makes the
workflow compatible with the broader Bioconductor ecosystem.

This vignette demonstrates four Bioconductor-oriented workflows:

- microbiome relative abundances stored in a `TreeSummarizedExperiment`,
- single-cell RNA-seq UMI counts stored in a `SingleCellExperiment`,
- single-cell RNA-seq non-UMI/read counts stored in a `SingleCellExperiment`,
- matched microbiome and metabolomics assays stored in a
  `MultiAssayExperiment`.

## Bioconductor Motivation

Submitting `Tweedieverse` to Bioconductor makes the package easier to install,
test, document, and reuse with established genomics infrastructure. The
Bioconductor submission model also encourages reproducible examples, formal
container classes, clear dependency management, and package checks across
operating systems. For users, this means that analyses can start from familiar
objects such as `SummarizedExperiment`, `SingleCellExperiment`,
`TreeSummarizedExperiment`, and `MultiAssayExperiment`, rather than from
unstructured tables whose orientation and metadata linkage must be inferred.

## Installation

After Bioconductor acceptance, the release version can be installed with:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("Tweedieverse")
```

The development version can be installed from GitHub:

```{r install-github, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("himelmallick/Tweedieverse")
```

```{r packages}

library(Tweedieverse)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(TreeSummarizedExperiment)
library(MultiAssayExperiment)

```

## Loading Packaged Example Data

The example containers used in this vignette are distributed with the package
under `inst/extdata`. This allows the vignette to run during Bioconductor
checks without network access.

```{r load-example-data}
load(system.file(
  "extdata",
  "tweedieverse_data.RData",
  package = "Tweedieverse"
))
```

## Microbiome Relative Abundance Example

For microbiome data, `Tweedieverse()` accepts a
`TreeSummarizedExperiment`. The assay contains taxa/features in rows and
samples in columns; `colData` contains sample metadata.

The example table contains relative abundances, so this analysis disables
additional offset normalization by setting `normalization = "NONE"`. For raw
microbiome counts, omit that argument to use the microbiome default `TSS`
size-factor offset.

```{r microbiome-fit}
microbiome_tse

microbiome_fit <- Tweedieverse(
  input_features = microbiome_tse,
  output = NULL,
  assay_name = "counts",
  domain = "microbiome",
  normalization = "NONE",
  fixed_effects = c("diagnosis", "antibiotics"),
  tweedie_p = NULL,
  median_comparison = FALSE,
  cores = 1
)

head(microbiome_fit)
```

## Single-Cell RNA-seq UMI Count Example

For single-cell RNA-seq data, `Tweedieverse()` accepts a
`SingleCellExperiment`. The UMI example uses integer molecule counts.

UMI counts are count-like. Here we pin the variance power to `p = 1`, the
Poisson variance case, and use the single-cell domain. `normalization = "SCRAN"`
computes size factors for the model offset without modifying the assay.

```{r umi-fit}
umi_sce

umi_fit <- Tweedieverse(
  input_features = umi_sce,
  output = NULL,
  assay_name = "counts",
  domain = "single_cell",
  normalization = "SCRAN",
  fixed_effects = c("condition", "batch"),
  tweedie_p = 1,
  median_comparison = FALSE,
  cores = 1
)

head(umi_fit)
```

## Single-Cell RNA-seq Non-UMI / Read Count Example

The non-UMI example uses read-count-like measurements. These can have a more
flexible mean-variance relationship than UMI counts, so we leave
`tweedie_p = NULL` and let Tweedieverse estimate the index after filtering.

```{r nonumi-fit}
nonumi_sce

nonumi_fit <- Tweedieverse(
  input_features = nonumi_sce,
  output = NULL,
  assay_name = "counts",
  domain = "single_cell",
  normalization = "SCRAN",
  fixed_effects = c("condition", "batch"),
  tweedie_p = NULL,
  median_comparison = FALSE,
  cores = 1
)

head(nonumi_fit)
```

## Multi-omics Container

For multi-omics data, `Tweedieverse()` accepts a `MultiAssayExperiment`.
The packaged Franzosa et al. example is stored as an `.RData` file in
`inst/extdata`. It contains matched microbiome species abundances and
metabolite profiles. Each omics layer is analyzed independently, and per-omics
parameters can be supplied as named vectors or lists.

```{r multiassay}
franzosa_example_mae

multi_fit <- Tweedieverse(
  input_features = franzosa_example_mae,
  output = NULL,
  assay_name = "abundance",
  domain = c(
    taxaAbundanceProfile = "microbiome",
    harmonizedMetaboliteProfile = "custom"
  ),
  normalization = c(
    taxaAbundanceProfile = "NONE",
    harmonizedMetaboliteProfile = "MEDIAN"
  ),
  fixed_effects = c("diseaseCat_clean", "Age"),
  tweedie_p = list(
    taxaAbundanceProfile = NULL,
    harmonizedMetaboliteProfile = NULL
  ),
  median_comparison = FALSE,
  max_significance = 1,
  cores = 1
)

names(multi_fit)
head(multi_fit$taxaAbundanceProfile)
head(multi_fit$harmonizedMetaboliteProfile)
```

## Session Info

```{r session-info}
sessionInfo()
```
